pyg4ometry: a python library for the creation of Monte Carlo particle transport geometries

### 2019-02-20 paper/pyg4ometry.tex
Added example

\author[a]{Stewart T. Boogert\corref{author}}
\author[a]{Andrey Abramov}
\author[a]{Laurence Nevay}
\author[a]{William Shields}

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called particle transport (PT) programmes, these 
include MCMP, FLUKA, Geant3 and Geant4. Each code either has a particular specialism or target user community.  PT Monte Carlo simulations have 
diverse uses, shielding calculations for radiological protection, detector performance, medical imaging and therapy, space radiation environment simulation 
are some examples. A fundamental requirement of all of the codes is to supply a computer readable description of the physical geometry that the particles 
are passing through.  The creation of geometry files is typically a very time consuming activity, the simulation validity and performance is directly dependent 
on the quality of the geometry. The requirement of a  geometry system that allows the conversion between files prepared for different codes will enable cross 
checks of the physics processes in particle transport codes. Parallel to the creation of geometry files for PT programs, usually computer assisted design (CAD) 
files exist for systems which need to be simulated in PT programmes. The fundamental geometric representations in CAD files is typically not amenable for PT 
programmes.  For these reasons it is advantageous to create a software tool that allows particle transport code users to rapidly develop geometry files, convert between 
common PT programme geometry formats and finally import CAD models. This paper describes such geometry creation and conversion package 
called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry description. The main features of  {\em pyg4ometry} 
are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry descriptions, conversion from CAD formats 
(STEP and IGES) based on FreeCAD-OpenCascade and powerful geometry visualisation tools based on VTK.

@@

\subsection{Geant4 geometery}
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes XX fundamental solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} 
and {\it Physical Volumes}. A Logical Volume is a volume that has an outer solid but also material, magnetic field and zero or more physical {\it daughter} volumes. A Physical 
Volume is a placement (or instance) of a Logical Volume.

@@

\subsection{Internal data representation} 
The internal data representation follows closely the structure of GDML. A registry class aggregates python ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.

@@

\subsection{Overlap detection}
All PT transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when encountering such a 
situation.  A key feature of {\it pyg4ometry} is the detection of potential overlaps efficiently, it does this by performing a CSG intersection between solid instances 
and determining if the resulting mesh is empty.

@@

\section{CAD to Geant4 conversion }
STEP and IGES files can be loaded into pyg4ometry, via an interface based on FreeCAD. FreeCAD is an open source CAD/CAM program, which in turn 
is based on OpenCASCADE. A STEP CAD file could be considered as a hierarchical tree of {\it part assemblies} and {\it part features}, where a part assembly is a 
collection of part features.

\section{Geant4 and FLUKA conversion}
FLUKA geometry is based upon a limited set of primitive volumes which can be combined using logical operations. Many FLUKA primitives are infinite in extent, so 
to emulate FLUKA geometry construction large but finite extent Geant4 primitive solids are used. Table XXX shows the mapping between FLUKA and Geant4 objects.

\section{Performance of converted geometry in Geant4}
Clearly for Geant4 geometry defined in {\it pyg4ometry} the performance is exactly the same as it the geometry was specified in C++ and linked against the Geant4
libraries. A comparison of the performance of converted CAD or FLUKA geometry compared with native Geant4 solids is required so a user can optimise the available
computing resources against the human resources (required to convert geometry).

>>>>

\author[a]{Stewart T. Boogert\corref{author}}
\author[a]{Andrey Abramov}
\author[a]{Laurence Nevay}
\author[a]{William Shields}
\author[a]{Stuart Walker}

@@

%% main text
\section{Introduction} \label{sec:introduction}
There are numerous different software codes to simulate the passage of particles through material, so called particle transport (PT) programmes, these 
include MCMP, FLUKA, Geant3 and Geant4. All these codes are based on the Monte Carlo technique but each code either has a particular specialism 
or target user community.  PT Monte Carlo simulations have diverse uses, shielding calculations for radiological protection, detector performance, medical 
imaging and therapy, space radiation environment simulation are some examples. A fundamental requirement of all of the codes is to supply a computer 
readable description of the physical geometry that the particles are passing through.  The creation of geometry files is typically a very time consuming activity, 
the simulation validity and performance is directly dependent on the quality of the geometry. The requirement of a  geometry system that allows the conversion 
between files prepared for different codes will enable cross checks of the physics processes in particle transport codes. Parallel to the creation of geometry files 
for PT programs, usually computer assisted design (CAD) files exist for systems which need to be simulated in PT programmes. The fundamental geometric 
representations in CAD files is typically not amenable for PT programmes.  For these reasons it is advantageous to create a software tool that allows particle 
transport code users to rapidly develop geometry files, convert between common PT programme geometry formats and finally import CAD models. This paper 
describes such geometry creation and conversion package called {\em pyg4ometry}, implemented in {\em Python} and internally based on the Geant4 geometry 
description. The main features of  {\em pyg4ometry} are a python scripting API to rapidly design parametrised geometry, conversion to and from  FLUKA geometry 
descriptions, conversion from CAD formats  (STEP and IGES) based on FreeCAD-OpenCascade and powerful geometry visualisation tools based on VTK. 
The origin of pyg4ometry was a set of utilities to prepare geometry for an accelerator beam line simulation based on Geant4 called BDSIM. Accelerator physicists 
like specialists in other areas need a tool to quickly simulate beam losses, and the subsequent interaction of the charged particle beam with the accelerator material.

@@

\subsection{Geant4 geometery}
Geant4 geometry description is a mixture of BREP, CSG and tessellated concepts. Geant4 includes XX fundamental solids, although it does not store a sense 
of topology present in traditional CAD BREP systems. One of the fundamental solids is a {\em TessellatedSolid} which can used to represent STL or PLY files. 
Geant4 also provides the ability to perform boolean operations on these primitive solids. The richest and most flexible geometry description is currently used by
Geant4. Not only do solid objects need to be defined but also placed in a world coordinate system. Geant4 has two concepts which facilitate this {\it Logical Volumes} 
and {\it Physical Volumes}. A Logical Volume is a region of space  that is defined by an outer solid but also other attributes like material, magnetic field and zero or 
more physical {\it daughter} volumes. A Physical Volume is a placement (or instance) of a Logical Volume.

@@

\subsection{Internal data representation} 
The internal data representation follows closely the structure of GDML. A \verb|Registry| class aggregates python ordered dictionaries that are  used to store the main 
elements of a GDML file. As a user specifies geometry the registry is updated. When a user is finished with the geometry, the registry is written to disk as a GDML file.

@@

\subsection{Overlap detection}
All PT transport codes can not handle spatial overlap between two geometric objects and likely are to have ill defined behaviour when tracking particles  
throughsuch a situation.  A key feature of {\it pyg4ometry} is the detection of potential overlaps efficiently, it does this by performing a CSG intersection between solid instances 
and determining if the resulting mesh is empty. Generally the overlap detection algorithm proceeds as

{\small 
\begin{verbatim}
Overlap detection
\end{verbatim}
}

For a logical volume with $n_{\rm daughters}$ physical volumes, assuming meshes for solids with average number of vertices $n$, clearly this algorithm has complexity $\sim \mathcal{O}(n^2 n_{\rm daughters})$. This might seem like a prohibitive computational cost, but worth it considering the potential waste if small overlaps are present in the final 
PT simulation. This algorithm clearly favours geometry descriptions which have a high degree of logical volume reuse. Due to the discrete nature of triangular meshes it is not possible to have perfect detection of overlaps, especially when curved surfaces are considered.

@@

An example of the VTK output for code listing XXX is shown in Figure~\ref{fig:rapidModellingExample}.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=10cm]{./diagrams/rapidModelling.pdf}
\caption{VTK visualisation output from code example XXX.}
\label{fig:rapidModellingExample}
\end{center}
\end{figure}
Significantly more complex geometries can be developed using a structure similar to that shown in code listing XXX.

\section{CAD to Geant4 conversion }
STEP and IGES files can be loaded into pyg4ometry, via an interface based on FreeCAD. FreeCAD is an open source CAD/CAM program, which in turn 
is based on OpenCASCADE. A STEP CAD file could be considered as a hierarchical tree of {\it part assemblies} and {\it part features}, where a part assembly is a 
collection of part features. A part feature can be used to create  a triangular mesh which can used to create a pyg4ometry tessellated solid. Assignment of materials 
and visualisation attributes.

\section{Geant4 and FLUKA conversion}
FLUKA geometry is based upon a limited set of primitive volumes which can be combined using logical operations. Many FLUKA primitives are infinite in extent, so 
to emulate FLUKA geometry construction large but finite extent Geant4 primitive solids are used. Table XXX shows the mapping between FLUKA and Geant4 objects.
An ANTLR based expression parser is used to load and comprehend the FLUKA input syntax, an example is shown in XXX. Each FLUKA body is matched to a Geant4 
solid and a FLUKA region is represented by a Geant4 CSG tree is formed using the pyg4ometry classes.

\begin{enumerate} 
\item infinite planes
\item disjoint unions 
\end{enumerate}

\section{Performance of converted geometry in Geant4}
Clearly for Geant4 geometry defined in {\it pyg4ometry} the performance is exactly the same as it the geometry was specified in C++ and linked against the Geant4
libraries. A comparison of the performance of converted CAD or FLUKA geometry compared with native Geant4 solids is required so a user can optimise the available
computing resources against the human resources (required to convert geometry). BDSIM is used to load the pyg4ometry generated files  and run basic simulations where 
high energy particles are
